Deciding Bitvector Termination with Program Synthesis

### 2014-06-24 papers/termination/paper.tex
revised intro

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders. Consequently, approaches based on Ramsey's theorem 
search for linear termination arguments.

@@

While the summary  in Figure~\ref{fig:handletable} supports our observation that the majority of existing termination analyses are designed for
linear programs and linear ranking functions, it also identifies another common assumption that is being made: the equivalence of bit-vector/float semantics and integer/real semantics. Thus, most of the techniques treat
bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

@@

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
checking the validity of a termination argument.

Given the level of generality that we aim for, we phrase the termination problem as a second order satisfaction problem (indeed, a program synthesis problem) (Section~\ref{}).
While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors influencing our method's performance and
find that our technique %is semantic in nature and 
does not dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
Moreover, we find that our technique does actually gain from the generality of the termination argument. 
After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
suggestively refer to as the \emph{synthesis fragment}:

@@

\section{Solving the Synthesis Formula}
We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
have a program synthesiser kicking around, we present the design of one synthesiser
that can be used to solve the problem.

>>>>

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders. 
This explain why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

@@

While the summary  in Figure~\ref{fig:handletable} supports the observation that the majority of existing termination analyses are designed for
linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
\emph{the equivalence of bit-vector/float semantics and integer/real semantics}. Thus, most of the techniques treat
bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument.

For this purpose, we phrase the termination problem as a second order satisfaction problem and
we identify a fragment of second-order logic that is expressive enough to capture our formulation %several variations of the halting problem 
(Section~\ref{sec:second.order}), which we call \emph{the synthesis fragment}. We propose a method 
for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
exist, it will be found. With respect to its performance, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
This means that our procedure is not directly dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
complexity of its minimal termination argument. We argue that it is natural to expect programmers to write programs that have relatively short termination 
proofs as this will make them understandable.

%% Our approach to proving termination has two distinct features over previous work. 
%% First,
%% Second, our runtime is dependant on the KC of the shortest termination proof rather than the structure of the loop, number of variables, number of lassos...

%% We show that the probability of
%% a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
%% Moreover, we find that our technique does actually gain from the generality of the termination argument. 
%% After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.

@@

\section{Termination as Second-Order Satisfaction} \label{sec:second.order}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
suggestively refer to as the \emph{synthesis fragment}:

@@

\section{Solving the Synthesis Formula} \label{sec:synthesis}
We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
have a program synthesiser kicking around, we present the design of one synthesiser
that can be used to solve the problem.
